---
title: "What's a Community College (Seriously)?"
author: "Will Doyle"
date: "4/12/2021"
output: html_document
---

```{r,echo=FALSE}
knitr::opts_chunk$set(echo=FALSE,message=FALSE,warning=FALSE,results = "hide")
```


```{r}
library(tidyverse)
library(here)
```


```{r}
inst<-read_csv(here("data/cleaned/institutions.csv"))
```


In an insightful  [New America piece](https://www.newamerica.org/education-policy/edcentral/a-different-approach-to-free-community-college/), Kevin Carey points out that a free community college plan depends crucially on the definition of community college.  Carey is right to point out is that we don't have a single definition of a community college. Many states don't have any colleges called "community colleges", and some states don't have any institutions that fulfill the classic missions of community colleges. 

Carey's piece includes a good breakdown of the ratio of associate's degrees to bachelor's degrees awarded by state. I was curious what percent and number of students in each state might be counted as "community college" students using different possible definitions of "community college."



```{r}
inst<-inst%>%
  ## create four different definitions of what a "cc is"
  ## IPEDS:sector, already defined
  ## Has CC in name
  mutate(`CC in Name`=ifelse(
    str_detect(instnm,"Community College") & control==1,1,0))%>%
  ## Carnegie Classification
  mutate(`Carnegie Associates Inst`=
           ifelse(c18ipug%in%c(1:3,5) &control==1,1,0)) %>%
  ## Less  than .5 of degrees are bachelors 
  mutate(all_degrees=Bachelors  +
         Masters +
         PhD +  
         `Cert> 1`+
         Postbac+ 
         Associates+
         `Cert<1`  )%>%
  mutate(prop_bach=Bachelors/all_degrees)%>%
  mutate(`Non-Bachelors`=ifelse(prop_bach<.5&control==1,1,0))%>%
  rename(State=name)
```


```{r}
## IPEDS Dataset
inst_ipeds<-inst%>%
  group_by(State)%>%
  mutate(`FTE in State`=sum(fteug))%>%
  group_by(State,sector)%>%
  summarize(`FTE in IPEDS Sector`=sum(fteug),
            `Total FTE`=mean(`FTE in State`))%>%
  mutate(`Percent in IPEDS Sector`=(`FTE in IPEDS Sector`/`Total FTE`)*100)%>%
  filter(sector==4)%>%
  select(State,`FTE in IPEDS Sector`,`Percent in IPEDS Sector`)
```

```{r}
## Carnegie
inst_carnegie<-inst%>%
  group_by(State)%>%
  mutate(`FTE in State`=sum(fteug))%>%
  group_by(State,`Carnegie Associates Inst`)%>%
  summarize(`FTE in Carnegie Associates`=sum(fteug),
            `Total FTE`=mean(`FTE in State`))%>%
  mutate(`Percent in Carnegie Associates`=
           (`FTE in Carnegie Associates`/`Total FTE`)*100)%>%
  filter(`Carnegie Associates Inst`==1)%>%
  select(State,`FTE in Carnegie Associates`,`Percent in Carnegie Associates`)
```


```{r}
## CC in Name
inst_cc_name<-inst%>%
  group_by(State)%>%
  mutate(`FTE in State`=sum(fteug))%>%
  group_by(State,`CC in Name`)%>%
  summarize(`FTE in Named CCs`=sum(fteug),
            `Total FTE`=mean(`FTE in State`))%>%
  mutate(`Percent in Named CCs`=
           (`FTE in Named CCs`/`Total FTE`)*100)%>%
  filter(`CC in Name`==1)%>%
  select(State,`FTE in Named CCs`,`Percent in Named CCs`)
```


```{r}
##  Non-Bachelors
inst_not_bach<-inst%>%
  group_by(State)%>%
  mutate(`FTE in State`=sum(fteug))%>%
  group_by(State,`Non-Bachelors`)%>%
  summarize(`FTE in Primarily Non-Bachelors`=sum(fteug),
            `Total FTE`=mean(`FTE in State`))%>%
  mutate(`Percent in Primarily Non-Bachelors`=
           (`FTE in Primarily Non-Bachelors`/`Total FTE`)*100)%>%
  filter(`Non-Bachelors`==1)%>%
  select(State,
         `FTE in Primarily Non-Bachelors`,
         `Percent in Primarily Non-Bachelors`,
         `Total FTE`)
```

```{r}
df<-inst_ipeds%>%
  full_join(inst_carnegie,by="State")%>%
  full_join(inst_cc_name,by="State")%>%
  full_join(inst_not_bach)%>%
  mutate(
    across(everything(), ~replace_na(.x, 0)) ## NAs to 0s
  )
```



## IPEDS definitions

Colleges and universities report data about themselves to the federal government through the Integrated Postsecondary Education Data System (IPEDS). IPEDS classifies institutions that award even 1 bachelor's degree per year as 4-year institutions. This misses a large number of institutions that either are community colleges or fulfill the roles of community colleges in their state. Here's the percent of students* in each state that attend "community colleges" if the definition is public two year institutions that don't award any bachelor's degrees:

## Percent of Students Enrolled in Public Two-Year Colleges, by State: 2019
```{r}
df%>%
  ggplot(aes(x=fct_reorder(State,`Percent in IPEDS Sector`),
                           y=`Percent in IPEDS Sector`))+
  theme_minimal()+
  geom_point(color="blue",alpha=.5)+
  coord_flip()+
  xlab("State")+
  ylab("Percent in  Sector")+
  theme(text = element_text(size = 10)) 
```

Source: IPEDS, 2019

This matches up with this [table](https://nces.ed.gov/programs/digest/d19/tables/dt19_304.80.asp) published by NCES in 2020. There are some surprises here: Florida, a national leader in the development of community  colleges, is reported as having about three percent of its enrollment in public two year colleges. Delaware and Alaska do not have any students enrolled in institutions that the federal government classified as public two-year institutions.  

## Carnegie  Classification-- Instructional Program: Associates Institutions

The [Carnegie Classification](https://carnegieclassifications.iu.edu/classification_descriptions/ugrad_program.php)  includes a definition of instructional programs that doesn't have a "community college" definition but does include "Associates Institutions" of different types. The graph below shows the percent of students enrolled in all types of Associates institutions compared with the percent enrolled in public two-year colleges as defined by NCES.

```{r}
df%>%
  ggplot(aes(x=fct_reorder(State,`Percent in IPEDS Sector`),
                           y=`Percent in IPEDS Sector`))+
  theme_minimal()+
  geom_point(color="blue",alpha=.5)+
  geom_point(data=df,
             aes(x=fct_reorder(State,`Percent in IPEDS Sector`),
                           y=`Percent in Carnegie Associates`),color="lightgreen")+
  coord_flip()+
  xlab("State")+
  ylab("Percent in Sector, Blue=Pub 2yr, Green=Carnegie Class")+
    theme(text = element_text(size = 10)) 
```

Source: IPEDS, 2019

Using the Carnegie definition there are considerably higher enrollments in Associates institutions as opposed to public two-year institutions. The biggest gains would come in the biggest states: Florida adds 280,000 students and California adds 144,000 students, but much smaller Washington adds 120,000 students. Overall, using the Carnegie Associates definition to designate an institution a "community college"  adds 916,000 students as opposed to using public two-year college.  


```{r}
df%>%
  mutate(diff=`FTE in IPEDS Sector`- `FTE in Carnegie Associates`)%>%
  arrange(diff)%>%
  select(State, diff)%>%
  print(n=50)


df%>%ungroup()%>%
  mutate(diff=`FTE in IPEDS Sector`- `FTE in Carnegie Associates`)%>%
  filter(diff<0)%>%
  summarize(sum(diff))


```


## Institutions with Community College in the Name

The simplest way to define an institution as a community college may be to use its name. The graph below compares enrollment in public two-year institutions with enrollment in institutions that include "community college" in their name. 
```{r}
df%>%
  ggplot(aes(x=fct_reorder(State,`Percent in IPEDS Sector`),
                           y=`Percent in IPEDS Sector`))+
  theme_minimal()+
  geom_point(color="blue",alpha=.5)+
  geom_point(data=df,
             aes(x=fct_reorder(State,`Percent in IPEDS Sector`),
                           y=`Percent in Named CCs`),color="orange")+
  coord_flip()+
  xlab("State")+
    ylab("Percent in Sector, Blue=Pub 2yr, Orange=Named Community Colleges")+
    theme(text = element_text(size = 10)) 
  
```

Source: IPEDS, 2019

Most states have considerably more public two-year colleges than they have institutions with the term "Community College' in their name. For instance, California has about 40 percent of full time equivalent students in public two years, but only about 10 percent of them attend institutions that call themselves "community colleges." Wisconsin, Georgia, Idaho all go to 0 because they don't have any institutions named "community" colleges. Limiting eligibility to institutions that have the name "community college" would mean 3.5 million fewer students with eligibility.   


```{r}
df%>%
  mutate(diff=`FTE in IPEDS Sector`- `Percent in Named CCs`)%>%
  arrange(diff)%>%
  select(State, diff)%>%
  print(n=50)


df%>%ungroup()%>%
  mutate(diff=`FTE in IPEDS Sector`- `Percent in Named CCs`)%>%
  filter(diff>0)%>%
  summarize(sum(diff,na.rm=TRUE))

```

## Proportion of Degrees

What if we used the proportion of degrees awarded? Sometimes the proportion of degrees that are associates is used, but I took another tack and instead defined it as institutions where less than half of all degrees awarded in a single year are bachelor's degrees. 

```{r}
df%>%
  ggplot(aes(x=fct_reorder(State,`Percent in IPEDS Sector`),
                           y=`Percent in IPEDS Sector`))+
  theme_minimal()+
  geom_point(color="blue",alpha=.5)+
  geom_point(data=df,
             aes(x=fct_reorder(State,`Percent in IPEDS Sector`),
                           y=`Percent in Primarily Non-Bachelors`),color="purple")+
  coord_flip()+
  xlab("State")+
    ylab("Percent in Sector, Blue=Pub 2yr, Purple=Non-Bachelor's Inst")+
    theme(text = element_text(size = 10)) 
```

Source: IPEDS, 2019

```{r}
df%>%
  mutate(diff=`FTE in IPEDS Sector`- `Percent in Primarily Non-Bachelors`)%>%
  arrange(diff)%>%
  select(State, diff)%>%
  print(n=50)


df%>%ungroup()%>%
  mutate(diff=`FTE in IPEDS Sector`- `Percent in Primarily Non-Bachelors`)%>%
  filter(diff>0)%>%
  summarize(sum(diff,na.rm=TRUE))

```
The biggest change here is in the very small (population-wise) state of Alaska, where most campuses mostly award degrees other than Bachelor's degrees. This definition would result in 3.5 million *more* students being in "community colleges" than if the public two-year definition were used.  

Which is the right definition? I don't know! It does seem to me that focusing on the first two years of public higher education, without reference to type of institution is a better way to go. The main point is that states are VERY different in the way that they structure their systems of higher education, and there are big implications for students for any arbitrary definition of "community college" that we come up with. 

* I define students as Full Time Equivalent (FTE) students, which basically means taking the total number of credit hours and dividing by 12. Headcount enrollment--the actual number of people attending-- is really important, but state-to-state comparisons aren't ideal using headcount because of variation in state policies supporting part-time students. The data is from [IPEDS, 2019](https://nces.ed.gov/ipeds/use-the-data). Code available: (https://github.com/wdoyle42/ccwhat).
